---
title: "NLP with Disaster Tweets - Model Comparison"
author: "Michael Swindle"
format:
  html:
    toc: true
    code-fold: true
  pdf:
    documentclass: scrartcl
execute:
  echo: true
  warning: false
  message: false
---

<!-- inline dark‑mode CSS -->
```{=html}
<style>
body            { background-color:#141A22; color:#ffffff; }
pre, code       { background-color:#1e2733; color:#ffffff; }
a               { color:#4da3ff; }
table, th, td   { border-color:#666666; }
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', dpi = 96)
library(tidyverse)
library(caret)
library(xgboost)
library(word2vec)
library(textstem)
library(tm)
source("helpers.R")
set.seed(123)

train_master <- read_csv("train.csv")
test_master  <- read_csv("test.csv")
```

## 1 Overview

Kaggle provides 10 000 labelled tweets and asks: *Is this tweet about a real disaster?*  
Three pipelines scale from sparse TF‑IDF counts to dense Word2Vec embeddings.

---

## 2 Pipelines

| Version | Feature set | Key steps |
|---------|-------------|-----------|
| **v7**  | TF‑IDF      | Lower‑case, punctuation strip, stop‑word removal, XGBoost |
| **v8**  | Clean TF‑IDF | Adds UK→US spelling map, extra noise removal |
| **v9**  | Word2Vec    | 100‑dim CBOW vectors averaged per tweet, XGBoost |

---

## 3 Results – Internal vs Kaggle

```{r compare, fig.cap="Internal 20 % hold‑out vs Kaggle leaderboard"}
# internal split
internal <- tibble(
  Pipeline = c("v7 TF‑IDF", "v8 Clean TF‑IDF", "v9 Word2Vec"),
  Internal = c(run_v7(train_master, test_master),
               run_v8(train_master, test_master),
               run_v9(train_master, test_master))
)

# hard‑coded leaderboard scores
kaggle <- tibble(
  Pipeline = c("v7 TF‑IDF", "v8 Clean TF‑IDF", "v9 Word2Vec"),
  Kaggle   = c(0.56114, 0.57370, 0.76953)
)

results <- internal %>% left_join(kaggle, by = "Pipeline")

# table
knitr::kable(results, digits = 3,
             col.names = c("Pipeline", "Internal F1", "Kaggle F1"))

# grouped bars
plot_df <- results %>% pivot_longer(-Pipeline, names_to = "Source", values_to = "F1")
ggplot(plot_df, aes(Pipeline, F1, fill = Source)) +
  geom_col(position = position_dodge(0.7), width = 0.6, colour = "black") +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Internal vs Kaggle F1 scores",
       y = "F1 score",
       fill = "Score source") +
  theme_minimal(base_size = 12) +
  theme(
    plot.background  = element_rect(fill = "white", colour = "white"),
    panel.background = element_rect(fill = "white", colour = "white")
  )
```

### Why Kaggle is lower

Random splits leak tweets from the same event into both train and test, inflating F1.  
Kaggle’s hidden set keeps events together and has a different class balance, so scores drop.

---

## 4 Lessons Learned

* Word2Vec gains about 0.20 F1 over sparse counts on both evaluations.  
* Extra cleaning offers negligible benefit.  
* Robust validation is essential – naïve random splits overstate performance.

---

## 5 Next Steps

1. Hyper‑parameter search.  
2. BERT baseline via HuggingFace.  
3. Ensemble TF‑IDF and embeddings.

---

## Appendix

<details>
<summary><strong>Full session info (click to expand)</strong></summary>

```{r session-info}
sessionInfo()
```

</details>
